---
title: "Analyzing Gerrymandering in Louisiana"
author: "Joseph Holler, Peter Kedron, Emily Zhou"
date: "`r Sys.Date()`"
output:
  html_document:
    css: styles.css
    code_folding: hide
    highlight: haddock
    theme: lumen
    toc: yes
    toc_depth: 4
    toc_float: yes
urlcolor: blue  
editor_options:
  markdown:
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<div class="date">
  Version 1.0 | First Created May 18, 2026 | Updated May 20, 2026
</div>

```{r sets up, warning = FALSE, message = FALSE, include=FALSE}
# global options 
options(scipen=999)
options(tigris_class = "sf")
options(digits = 3)

# list of required packages
packages <- c("tidyverse",
              "dplyr",
              "here",
              "ggplot2",
              "lwgeom",
              "tidycensus",
              "knitr", 
              "htmltools",
              "kableExtra",
              "markdown",
              "ggcorrplot",
              "gridExtra",
              "sf")

# install and load required packages
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE, quietly=TRUE)
      library(x, character.only = TRUE)
    }
  }
)

theme1 <- theme(plot.subtitle = element_text(size = 12,face = "italic"),
                legend.position = "none",
                plot.title = element_text(size = 14, face = "bold"), 
                axis.text.x=element_text(size=8),
                axis.text.y=element_text(size=8), 
                axis.title=element_blank(), 
                panel.background = element_blank(),
                panel.border = element_rect(colour = "grey", fill=NA, linewidth =0.8))
palette1 <- c("#6EA6C3", "#99CFAB", "#FEE59C","#DC6D85", "#FFD190", "#F1B6AC", "#F7B9C8", "#C1DDF4","#C1E3CD", "#ECDCA8" ) 
green_palette <- c("#f7fcf5",  "#C1E3CD",  "#99CFAB",  "#41ab5d",  "#006d2c")

theme2 <- theme(
    plot.subtitle  = element_text(size = 9, face = "italic"),
    plot.title     = element_text(size = 12, face = "bold"),
    axis.text.x    = element_text(size = 9),
    axis.text.y    = element_text(size = 9),
    axis.title     = element_text(size = 9))
  
theme3 <- theme_bw() +
  theme(plot.subtitle = element_text(size = 12, face = "italic"),
        plot.title = element_text(size = 14, face = "bold"),
        axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        axis.title = element_blank(),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "grey", fill = NA, linewidth = 0.8))

```

# 1. Introduction



# 2. Data and Variables

```{r load data, warning = FALSE, message=FALSE, results='hide'}

# file paths
districts_file <- here("data", "raw", "public", "Act_2_1st_ES_(2024)_-_Congress.shp")
blockgroup_file <- here("data", "raw", "public", "LA_block_groups.gpkg")
precincts_file <- here("data", "derived", "public", "LA_precincts.gpkg")
census_metadata_file <- here::here("data", "metadata", "census2020pl_vars.csv")

# read in the data
districts <- st_read(districts_file)
precincts <- st_read(precincts_file, layer = "LA_precincts_clipped")
blockgroups <- st_read(blockgroup_file)
census2020pl_vars <- read.csv(census_metadata_file)

```


```{r plot data, fig.height=6, fig.width=12, dpi=300, echo=FALSE, warning=FALSE, message=FALSE}
# district
dst <- ggplot() +
  geom_sf(data = districts, fill = "#C1E3CD", col = "#006d2c", alpha = 0.5) +
  theme1 +
  labs(title = "2021 Louisiana Congressional Districts")
# precincts
prct <- ggplot() +
  geom_sf(data = precincts, fill =  "#C1DDF4", col = "#6EA6C3", alpha = 0.5) +
  theme1 +
  labs(title = "2020 Louisiana Precincts")
# block groups
bg <- ggplot() +
  geom_sf(data = blockgroups, fill = "#F1B6AC", col = "#DC6D85", alpha = 0.5) +
  theme1 +
  labs(title = "2020 Louisiana Block Groups")

grid.arrange(dst, prct, bg, nrow = 1)
```


# 3. Data Processing

## 3.1 Spatial Transformations

```{r reproject data, warning=FALSE, message=FALSE}
districts <- districts |> st_transform(4269) 
st_geometry(districts) <- "geom"

blockgroups <- blockgroups |> st_transform(4269)
st_geometry(blockgroups) <- "geom"

precincts <- precincts |> st_make_valid() |> st_transform(4269) |> 
st_geometry(precincts) <- "geom"

```

Census data needs to be transformed from the WGS 1984 geographic coordinate system to the NAD 1983 geographic coordinate system.


## 3.2 Compute  Variable of Interest

Find the total of people identifying as Black or African American as one race or any combination of multiple races.
First, make a list of all the variables inclusive of people identifying as Black or African American.

```{r find variables, warning=FALSE, message=FALSE}
minority_cat <- "Black"
minority_vars <- census2020pl_vars |> 
  dplyr::filter(str_detect(name, "P3"),
                str_detect(label, minority_cat)) |> 
  select(-concept)
minority_vars |> 
  kable(format = "html") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
```

Next, calculate new columns.
`Black` is a sum of all 32 columns shown above, in which any of the racial categories by which someone identifies is Black or African American.  
`Total` is a copy of the population 18 years or over, variable `P3_001N`.  
`pctMinority` is calculated as `Black / Total * 100`  
`CheckPct` is calculated as the percentage of the population 18 years or older that is either white of one race only (`P3_003N`) *or* Black or African American as calculated above. 

```{r calculate variables, warning=FALSE, message=FALSE}
blockgroups_calc <- blockgroups |> 
  rowwise() |> 
  mutate(minority = sum(c_across(all_of(minority_vars$name)))) |> 
  ungroup() |> 
  mutate(sarea = st_area(geom),
         Total = P3_001N,
         pctMinority = minority / Total * 100,
         CheckPct = (minority + P3_003N) / Total * 100) |> 
  select(GEOID, sarea, minority, Total, pctMinority, CheckPct)
```

```{r calculate variables for precincts, warning=FALSE, message=FALSE}
precincts <- precincts |> 
  mutate(
  totV = G24PRERTRU + G24PREDHAR,
  mpV = G24PREDHAR) |>
  mutate(sarea = st_area(geom))
```

Map the percentage of the population 18 or over that is a minority.

```{r map, message=FALSE, warning=FALSE, fig.height=8, fig.width=12, dpi=300, echo=FALSE}

bg_M <- ggplot() +
  geom_sf(data = blockgroups_calc, aes(fill = pctMinority), col = "grey90", lwd = 0.1) +
  scale_fill_gradient(low = "#f7fcf5", high = "#006d2c", name = "Pct Minority") +
  theme1 +
  labs(title = "Minority Population (18+) By Block Group", fill = "Percentage") + 
  theme(
    legend.position = "bottom",
    axis.title = element_text(size = 9))

v_M <- ggplot() +
  geom_sf(data = precincts, aes(fill = mpV / totV * 100), col = "grey90", lwd = 0.1) +
  #blue palette
  scale_fill_gradient(low = "#F0FFFF", high = "#6EA6C3", name = "Pct Minority Voters") +
  theme1 +
  labs(title = "Minority Party Vote By Precinct", ll = "Percentage") + 
  theme(
    legend.position = "bottom",
    axis.title = element_text(size = 9))

grid.arrange(bg_M, v_M, nrow = 1)
```


# 4. Analysis

## 4.1 Area Weighted Reaggregation

Estimate the white and black voting age populations using AWR with block groups.
Why do this when POPULATION, BLACK, and WHITE variables are already in the table?
First, this is the total population, but we should care more about the voting age population.
Second, we may want to categorize and calculate `BLACK` differently from the state of Alabama.

```{r bg awr, message=FALSE, warning=FALSE}

district_estimates <- st_intersection(blockgroups_calc, districts) |> 
  st_make_valid() |> 
  mutate(
    awTot = Total * as.numeric(st_area(geom) / sarea),
    awMinor = minority * as.numeric(st_area(geom) / sarea)
  ) |> 
  st_drop_geometry() |> 
  group_by(DISTRICT) |> 
  summarize(bgTotal = sum(awTot),
            bgMinority = sum(awMinor))

districts_join <- districts |> 
  left_join(district_estimates, by = "DISTRICT") |> 
  mutate(pctMinoritybg = round(bgMinority / bgTotal * 100, 1))
```

```{r voting awr, warning=FALSE, message=FALSE}
district_estimates <- st_intersection(precincts, districts) |> 
  st_make_valid() |> 
  mutate(
    awTot = totV * as.numeric(st_area(geom) / sarea),
    awMinor = mpV * as.numeric(st_area(geom) / sarea)
  ) |> 
  st_drop_geometry() |> 
  group_by(DISTRICT) |> 
  summarize(totV = sum(awTot),
            mpV = sum(awMinor))

districts_join <- districts_join |> 
  left_join(district_estimates, by = "DISTRICT") |> 
  mutate(pctMP = round(mpV / totV * 100, 1))
```


```{r print district table, warning=FALSE, message=FALSE, echo=FALSE}
districts_join |> st_drop_geometry() |> 
  dplyr::select(DISTRICT, bgTotal, bgMinority, pctMinoritybg, totV, mpV, pctMP) |>
  rename(
    "District" = DISTRICT,
    "Total Pop" = bgTotal,
    "Minority Pop" = bgMinority,
    "Pct Minority" = pctMinoritybg,
    "Total Voters" = totV,
    "Minority Party Votes" = mpV,
    "Pct Minority Party Votes" = pctMP
  ) |>
  kable() |> 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
```

## 4.2 Convex Hulls

```{r make convex hulls, warning=FALSE, message=FALSE, dpi=300, fig.height=8, fig.width=12}
d_ch <- districts |> 
  st_convex_hull() |> 
  mutate(chArea = st_area(geom)) # compute the area of the convex hull

ggplot() +
  geom_sf(data = districts, fill = "#C1E3CD", col = "#006d2c", alpha = 0.8, linewidth = 0.2) +
  geom_sf(data = d_ch, fill = "#F0FFFF", col = "#6EA6C3", alpha = 0.5, linewidth = 0.6) +
  theme1 +
  labs(title = "Districts and Their Convex Hulls")
```

```{r convex hull pop estimates, warning=FALSE, message=FALSE}
# estimate the population in the convex hulls using AWR with block groups
d_ch_est <- blockgroups_calc |> 
  st_intersection(d_ch) |> 
  st_make_valid() |> 
  mutate(
    fArea = st_area(geom),
    awTot = Total * as.numeric(fArea / sarea),
    awMinor = minority * as.numeric(fArea / sarea),
  ) |> 
  st_drop_geometry() |> 
  group_by(DISTRICT, chArea) |> 
  summarize(chTotal = sum(awTot),
            chMinority = sum(awMinor))
```


```{r print convex hull table, warning=FALSE, message=FALSE, echo=FALSE}
d_ch_est |>
  dplyr::select(DISTRICT, chTotal, chMinority) |>
  rename(
    "District" = DISTRICT,
    "Total Pop in Convex Hull" = chTotal,
    "Minority Pop in Convex Hull" = chMinority
  ) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
```


```{r convex hull voting estimates, warning=FALSE, message=FALSE}
# voting population estimates in the convex hulls using AWR with precincts
d_ch_estV <- precincts |> 
  st_intersection(d_ch) |> 
  st_make_valid() |> 
  mutate(
    fArea = st_area(geom),
    awTot = totV * as.numeric(fArea / sarea),
    awMinor = mpV * as.numeric(fArea / sarea),
  ) |> 
  st_drop_geometry() |> 
  group_by(DISTRICT) |> 
  summarize(chtotV = sum(awTot),
            chmpV = sum(awMinor))
```


```{r print convex hull voting table, warning=FALSE, message=FALSE, echo=FALSE}
d_ch_estV |>
  dplyr::select(DISTRICT, chtotV, chmpV) |>
  rename(
    "District" = DISTRICT,
    "Total Voters in Convex Hull" = chtotV,
    "Minority Party Votes in Convex Hull" = chmpV
  ) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
```

## 4.3 Minimum Bounding Circle

```{r make mbc, warning=FALSE, message=FALSE, dpi=300, fig.height=8, fig.width=12}
d_mbc <- districts |> 
  lwgeom::st_minimum_bounding_circle() |> 
  mutate(mbcArea = st_area(geom))

ggplot() +
  geom_sf(data = districts, fill = "#C1E3CD", col = "#006d2c", alpha = 0.8, linewidth = 0.2) +
  geom_sf(data = d_mbc, fill = "#F0FFFF", col = "#6EA6C3", alpha = 0.5, linewidth = 0.6) +
  theme1 +
  labs(title = "Districts and Their Minimum Bounding Circles")
```

```{r mbc pop estimates, warning=FALSE, message=FALSE}
d_mbc_est <- blockgroups_calc |> 
  st_intersection(d_mbc) |> 
  st_make_valid() |> 
  mutate(
    fArea = st_area(geom),
    awTot = Total * as.numeric(fArea / sarea),
    awMinority = minority * as.numeric(fArea / sarea),
  ) |> 
  st_drop_geometry() |> 
  group_by(DISTRICT, mbcArea) |> 
  summarize(mbcTotal = sum(awTot),
            mbcMinority = sum(awMinority))
```


```{r print mbc table, warning=FALSE, message=FALSE, echo=FALSE}
d_mbc_est |>
  dplyr::select(DISTRICT, mbcTotal, mbcMinority) |>
  rename(
    "District" = DISTRICT,
    "Total Pop in MBC" = mbcTotal,
    "Minority Pop in MBC" = mbcMinority
  ) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
```

```{r mbc voting estimates, warning=FALSE, message=FALSE}
d_mbc_estV <- precincts |> 
  st_intersection(d_mbc) |> 
  st_make_valid() |> 
  mutate(
    fArea = st_area(geom),
    awTot = totV * as.numeric(fArea / sarea),
    awMinor = mpV * as.numeric(fArea / sarea),
  ) |> 
  st_drop_geometry() |> 
  group_by(DISTRICT) |> 
  summarize(mbctotV = sum(awTot),
            mbcmpV = sum(awMinor))
```


```{r print mbc voting table, warning=FALSE, message=FALSE, echo=FALSE}
d_mbc_estV |>
  dplyr::select(DISTRICT, mbctotV, mbcmpV) |>
  rename(
    "District" = DISTRICT,
    "Total Voters in MBC" = mbctotV,
    "Minority Party Votes in MBC" = mbcmpV
  ) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
```

## 4.4 Compactness and Composition Comparison

Join convex hull estimates to Districts with block group estimates.

Calculate compactness scores based on:

- the area and perimeter
- the area and the area of the convex hull
- the area and the area of the minimum bounding circle

“How different would the demographic and voting composition be if districts were represented by alternative geometries (convex hulls or minimum bounding circles), and how compact are those district shapes?”

Traditional shape compactness measures

```{r join results, warning=FALSE, message=FALSE}
districts_results <- districts_join |> 
  left_join(d_ch_est, by = "DISTRICT") |> 
  mutate(chPct = round(chMinority / chTotal * 100, 1),
         chDiff = pctMinoritybg - chPct,
         chAbsDiff = abs(chDiff)) |> 
  left_join(d_mbc_est, by = "DISTRICT") |> 
  mutate(mbcPct = round(mbcMinority / mbcTotal * 100, 1),
         mbcDiff = pctMinoritybg - mbcPct,
         mbcAbsDiff = abs(mbcDiff))  |> 
  left_join(d_ch_estV, by = "DISTRICT") |> 
  mutate(chPctV = round(chmpV / chtotV * 100, 1),
         chDiffV = pctMP - chPctV,
         chAbsDiffV = abs(chDiffV)) |> 
  left_join(d_mbc_estV, by = "DISTRICT") |> 
  mutate(mbcPctV = round(mbcmpV / mbctotV * 100, 1),
         mbcDiffV = pctMP - mbcPctV,
         mbcAbsDiffV = abs(mbcDiffV)) |> 
  # calculate compactness scores
  mutate(
    darea = st_area(geom),
    dperim = st_length(st_cast(geom, "MULTILINESTRING")),
    compact_shp = round( as.numeric((4 * pi * darea) / dperim^2), 2),
    compact_hull = round( as.numeric(darea / chArea), 2),
    compact_circ = round( as.numeric(darea / mbcArea), 2)
  )
```


```{r print comparison table, warning=FALSE, message=FALSE, echo=FALSE}
districts_results |> st_drop_geometry() |> 
  dplyr::select(DISTRICT, pctMinoritybg, chPct, chDiff, chAbsDiff, mbcPct, mbcDiff, mbcAbsDiff) |>
  rename(
    "District" = DISTRICT,
    "Pct Minority in District" = pctMinoritybg,
    "Pct Minority in Convex Hull" = chPct,
    "Difference (District - Convex Hull)" = chDiff,
    "Absolute Difference (District - Convex Hull)" = chAbsDiff,
    "Pct Minority in MBC" = mbcPct,
    "Difference (District - MBC)" = mbcDiff,
    "Absolute Difference (District - MBC)" = mbcAbsDiff
  ) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE)
```

# 5. Comparison

```{r correlation matrix, warning=FALSE, message=FALSE, fig.height=10, fig.width=12, dpi=300}
districts_results_cor <- districts_results |>
  st_drop_geometry() |>
  select(
    pctMinoritybg,
    chAbsDiff,
    mbcAbsDiff,
    compact_shp,
    compact_hull,
    compact_circ)

popcor <- ggcorrplot(
  round(cor(districts_results_cor), 1),
  colors = c("#6EA6C3", "#F7F0F5", "#DC6D85"),  
  type = "lower",
  insig = "blank",
  lab = TRUE,       
  lab_size = 3,     
  tl.cex = 8) +
  labs(
    title = "District Shape and Minority Population Composition") +
  theme2


districts_results_cor2 <- districts_results |> 
  st_drop_geometry() |> 
  select(pctMP,
         chAbsDiffV,
         mbcAbsDiffV,
         compact_shp,
         compact_hull,
         compact_circ)

votecor <- ggcorrplot(
  round(cor(districts_results_cor2), 1),
  colors = c("#6EA6C3", "#F7F0F5", "#DC6D85"),  
  type = "lower",
  insig = "blank",
  lab = TRUE,       
  lab_size = 3,      
  tl.cex = 8) +
  labs(
    title = "District Shape and Minority Voting Composition") +
  theme2

grid.arrange(popcor, votecor, nrow = 1)
```


This means:

actual district minority percentage minus compact-shape minority percentage.

So:

positive value = the actual district contains more minority population than its compact comparison shape
possible packing
negative value = the actual district contains less minority population than its compact comparison shape
possible cracking/dilution
near zero = the weird shape may not meaningfully affect composition                              


```{r compactness comparison plot, warning=FALSE, message=FALSE, fig.height=6, fig.width=12, dpi=300}
districts_results |>
  st_drop_geometry() |>
  select(DISTRICT, compact_shp, compact_hull, compact_circ) |>
  pivot_longer(cols = c(compact_shp, compact_hull, compact_circ),
               names_to = "measure", values_to = "score") |>
  mutate(
    measure = recode(measure,
      compact_shp  = "Polsby-Popper",
      compact_hull = "Convex Hull Ratio",
      compact_circ = "Reock (Circle Ratio)"),
    DISTRICT = factor(DISTRICT) |> fct_reorder(score)
  ) |>
  ggplot(aes(x = DISTRICT, y = score, fill = measure)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("#6EA6C3", "#99CFAB", "#DC6D85")) +
  labs(
    title = "Compactness Scores by District and Measure",
    subtitle = "Higher scores indicate more compact districts",
    x = "District", y = "Compactness Score", fill = "Measure"  ) +
  theme1 +
  theme(
    legend.position = "bottom",
    axis.title = element_text(size = 9))

```


```{r fig.height=12, fig.width=12, dpi=300}
# combine both into one long dataframe
combined <- bind_rows(
  districts_results |>
    st_drop_geometry() |>
    select(DISTRICT, actual = pctMinoritybg, ch = chPct, mbc = mbcPct) |>
    mutate(type = "Minority Population (%)"),
  districts_results |>
    st_drop_geometry() |>
    select(DISTRICT, actual = pctMP, ch = chPctV, mbc = mbcPctV) |>
    mutate(type = "Minority Party Vote (%)")
) |>
  mutate(
    DISTRICT = factor(DISTRICT),
    type = factor(type, levels = c("Minority Population (%)", "Minority Party Vote (%)"))
  )

ggplot(combined, aes(y = type)) +
  geom_segment(aes(x = ch, xend = mbc, yend = type),
               color = "darkgrey", linewidth = 1.5) +
  geom_point(aes(x = ch), color = "#99CFAB", size = 4) +
  geom_point(aes(x = mbc), color = "#DC6D85", size = 4) +
  geom_point(aes(x = actual),color ="#6EA6C3", size = 4) +
  facet_wrap(~ DISTRICT, ncol = 1, labeller = label_both) +
  labs(
    title = "Demographic Estimates by District and Geometry Type",
    subtitle = "Blue = actual district  |  Green = convex hull  |  Pink = min. bounding circle",
    x = "Percentage (%)", y = NULL) +
  theme3 +
  theme(
    strip.text       = element_text(face = "bold", size = 10),
    panel.spacing    = unit(1, "lines")
  )
```

