---
title: "Section 6: Comparison"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 6. Comparison

The last figure examines the statistical relationship between compactness, minority composition, and representational differences. The strongest pattern is the large negative correlations between: minority composition and compactness measures. This means that higher minority population percentages are associated with lower compactness scores and higher minority party vote shares are also associated with lower compactness. 

> Note: This pattern is consistent with the idea of packing, where districts with higher minority populations are drawn in more irregular shapes to concentrate those voters into fewer districts. 

Please consider: 

- Which compactness measure appears most strongly related to demographic change?

- Are compactness and representation strongly correlated?

- Do racial and partisan composition behave similarly?


```{r correlation matrix, warning=FALSE, message=FALSE, echo=FALSE, fig.height=6, fig.width=12, dpi=300}
districts_results_cor <- districts_results |>
  st_drop_geometry() |>
  mutate(chAbsDiff = 100 - chAbsDiff,
         mbcAbsDiff = 100 - mbcAbsDiff) |> 
  select(
    #pctMinoritybg,
    chAbsDiff,
    mbcAbsDiff,
    compact_shp,
    compact_hull,
    compact_circ)

popcor <- ggcorrplot(
  round(cor(districts_results_cor), 1),
  colors = c("#6EA6C3", "#F7F0F5", "#DC6D85"),  
  type = "lower",
  insig = "blank",
  lab = TRUE,       
  lab_size = 3,     
  tl.cex = 8) +
  labs(
    title = "District Shape and Minority Population Composition") +
  theme2


districts_results_cor2 <- districts_results |> 
  st_drop_geometry() |> 
  mutate(chAbsDiffV = 100 - chAbsDiffV,
         mbcAbsDiffV = 100 - mbcAbsDiffV) |> 
  select(#pctMP,
         chAbsDiffV,
         mbcAbsDiffV,
         compact_shp,
         compact_hull,
         compact_circ)

votecor <- ggcorrplot(
  round(cor(districts_results_cor2), 1),
  colors = c("#6EA6C3", "#F7F0F5", "#DC6D85"),  
  type = "lower",
  insig = "blank",
  lab = TRUE,       
  lab_size = 3,      
  tl.cex = 8) +
  labs(
    title = "District Shape and Minority Voting Composition") +
  theme2

grid.arrange(popcor, votecor, nrow = 1)
```


<span style="color: #6f7f6c; font-weight: bold;">Figure 7: Correlation Analysis</span>
